
---
output: github_document
---


# Giotto 0.3.1

- (optional) automatic installation of python modules through reticulate:  
  - provide your preferred python path
  - install giotto environment (automatic)
  - if you do not provide the python path and do not choose to install the giotto environment, then it will take the default python path  
- several bug fixes
- several mini-datasets are now included with Giotto for quick testing, example:  

```{r, eval=FALSE}
small_seqfish_expr_matrix = read.table(system.file("extdata", "seqfish_field_expr.txt", package = 'Giotto'))
small_seqfish_locations = read.table(system.file("extdata", "seqfish_field_locs.txt", package = 'Giotto'))
```

# Giotto 0.3.0

- Default spatial network created with **createSpatialNetwork** is now a Delaunay spatial network.
```{r, eval=FALSE}
# to create the old default kNN spatial network use:
createSpatialKNNnetwork(gobject)

# or use this function with the following setting
createSpatialNetwork(gobject, method = 'kNN')
```

- The function names for extracting spatial genes have changed:
```{r, eval=FALSE}
# binGetSpatialGenes is now:
binSpect(gobject) # binary Spatial extraction

# spatial_genes_python is now:
silhouetteRank(gobject)
```

- Fixed multiple bugs
- Improved speed by changing code to Rcpp and implementing parallelization options
- updated [howto's](./docs/getting_started.html) tutorials in Start section
- Finished the analysis of 10 different spatial datasets (tutorials are a work-in-progress)

# Giotto 0.2.4

- New examples on mouse kidney and brain using the recently released [10X Visium datasets](https://www.10xgenomics.com/spatial-transcriptomics/) (**NEW**)
- Added tools to identify spatial enrichment based on cell-type specific gene signature lists  (**NEW**)


# Giotto 0.2.3

- New example with 3D-like spatial data of the mouse hypothalamic preoptic region using [merFISH](https://science.sciencemag.org/content/362/6416/eaau5324) (**NEW**)  
- New example with 3D spatial data [STARmap](https://science.sciencemag.org/content/361/6400/eaat5691)
- New example with the highly sensitive data from [osmFISH](https://www.nature.com/articles/s41592-018-0175-z)
- New example on the Cerebellum with the scalable data from [Slideseq](https://science.sciencemag.org/content/363/6434/1463)
- New example on mouse olfactory bulb using immobilized primers on glass slides from [Spatial Transcriptomics](https://science.sciencemag.org/content/353/6294/78)
- Updated seqFISH+ cortex example (**NEW**)
- Updated STARmap cortex example (**NEW**)


# Giotto 0.2.2

- Implemented [SpatialDE](https://github.com/Teichlab/SpatialDE) and [trendsceek](https://github.com/edsgard/trendsceek)
- Updated support for 3D spatial data  
- Added support for the use of global instructions and automatically saving your plots (**NEW**)
- Add wrapper for differential expression with [MAST](https://github.com/RGLab/MAST) and [SCRAN](https://bioconductor.org/packages/release/bioc/html/scran.html)


