---
title: "Giotto Ecosystem"
date: 4/12/23
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Giotto Ecosystem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# 1. Explanation

We plan to curate an ecosystem of different, related packages to modularize Giotto Suite as we extend its functionalities. Here, we detail some different helper module(s) within the Giotto Ecosystem.

# 2. GiottoData

This package depends on Giotto Suite and leverages various functions within it for saving and loading Giotto objects. It contains dataset helper functions `loadGiottoMini()` and `getSpatialDataset()`. **Note that** `getSpatialDataset()` **was moved from Giotto Suite to GiottoData!**

In addition to dataset helpers, we have created mini Giotto Objects for testing Giotto Suite's capabilities rapidly and streamlining the tutorial experience. GiottoData currently includes two mini Giotto Objects which are derived from Vizgen and Visium data; more mini Giotto Objects will be published in the future. Further, we include mini objects, S4 subobjects of a Giotto Object (i.e. exprObj), in an effort to make the Giotto Object structure as transparent as possible.

GiottoData may be downloaded in a similar fashion to Giotto Suite:


```{r, eval=FALSE}
devtools::install_github("drieslab/GiottoData")
```


The scripts used to create both the mini giotto objects and mini objects are available in the repository so that we may extend and/or improve them and that you may utilize them for your own purposes!

## 2.1 Start Giotto


```{r, eval=FALSE}
# Ensure Giotto Suite is installed
if(!"Giotto" %in% installed.packages()) {
  devtools::install_github("drieslab/Giotto@suite")
}

library(Giotto)

# Ensure Giotto Data is installed
if(!"GiottoData" %in% installed.packages()) {
  devtools::install_github("drieslab/GiottoData")
}

library(GiottoData)

# Ensure the Python environment for Giotto has been installed
genv_exists = checkGiottoEnvironment()
if(!genv_exists){
  # The following command need only be run once to install the Giotto environment
  installGiottoEnvironment()
}
```


## 2.2 Create a Giotto object


```{r, eval=FALSE}
visium_mini = loadGiottoMini(dataset = 'visium')
```


# 3. Session Info


```{r, eval=FALSE}
sessionInfo()
```

```{r, eval=FALSE}
 
    R version 4.2.2 (2022-10-31 ucrt)
    Platform: x86_64-w64-mingw32/x64 (64-bit)
    Running under: Windows 10 x64 (build 22621)

    Matrix products: default

    locale:
    [1] LC_COLLATE=English_United States.utf8 
    [2] LC_CTYPE=English_United States.utf8   
    [3] LC_MONETARY=English_United States.utf8
    [4] LC_NUMERIC=C                          
    [5] LC_TIME=English_United States.utf8    

    attached base packages:
    [1] stats     graphics  grDevices utils     datasets  methods   base     

    other attached packages:
    [1] GiottoData_0.1.0 Giotto_3.2.1    

    loaded via a namespace (and not attached):
     [1] Rcpp_1.0.10       pillar_1.9.0      compiler_4.2.2    tools_4.2.2      
     [5] digest_0.6.30     jsonlite_1.8.3    evaluate_0.20     lifecycle_1.0.3  
     [9] tibble_3.2.1      gtable_0.3.3      lattice_0.20-45   png_0.1-7        
    [13] pkgconfig_2.0.3   rlang_1.1.0       Matrix_1.5-1      cli_3.4.1        
    [17] rstudioapi_0.14   parallel_4.2.2    yaml_2.3.7        xfun_0.38        
    [21] fastmap_1.1.0     terra_1.7-18      dplyr_1.1.1       knitr_1.42       
    [25] rappdirs_0.3.3    generics_0.1.3    vctrs_0.6.1       grid_4.2.2       
    [29] tidyselect_1.2.0  reticulate_1.26   glue_1.6.2        data.table_1.14.6
    [33] R6_2.5.1          fansi_1.0.4       rmarkdown_2.21    ggplot2_3.4.1    
    [37] magrittr_2.0.3    scales_1.2.1      codetools_0.2-18  htmltools_0.5.4  
    [41] colorspace_2.1-0  utf8_1.2.3        munsell_0.5.0    

```
