---
title: "Solutions to Installation Errors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Solutions to Installation Errors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Errors on MacOS

## Packages not Found

In the event that packages are inaccessible in the default installation of the Giotto miniconda environment, one troubleshooting method is provided here.


```{r, eval=FALSE}
# Restart the R session, while maintaining workspace variables.
# If using RStudio, the following command will do exactly that:
.rs.restartR()
# Direct reticulate to use Python within the Giotto Environment
reticulate::use_python(default_python_path)
# Check if packages exist again. Ensure function from above code block is defined.
missing_packages <- pkg_check()
retry_install <- length(missing_packages) > 0
if(retry_install){

  # Attempt to reinstall all packages.
  pkgs_w_versions <- c('pandas==1.5.1',
                       'networkx==2.8.8',
                       'python-igraph==0.10.2',
                       'leidenalg==0.9.0',
                       'python-louvain==0.16',
                       'python.app==1.4',
                       'scikit-learn==1.1.3')

  python_version = "3.10.2"

  py_pkgs = c('pandas','networkx',
              'igraph', 'leidenalg',
              'python-louvain','scikit-learn','python.app')

  if(Sys.info()[['sysname']] != "Darwin"){
    pkgs_w_versions = pkgs_w_versions[!grepl(pattern = 'python.app', x = pkgs_w_versions)]
    py_pkgs = py_pkgs[!grepl(pattern = 'python.app', x = py_pkgs)]
  }

  env_location <- reticulate::py_discover_config()$pythonhome
  partial_path_to_conda <- paste0(reticulate::miniconda_path(),'/envs/giotto_env')

  py_lou = pkgs_w_versions[grepl(pattern = 'python-louvain',x = pkgs_w_versions)]
  pip_packages = c("smfishhmrf", py_lou)
  pkgs_w_versions = pkgs_w_versions[!grepl(pattern = 'python-louvain',x = pkgs_w_versions)]

  if(.Platform[['OS.type']] == 'unix') {

    conda_full_path = paste0(partial_conda_path,'/','bin/conda')

    # Remove all previous installations
    reticulate::conda_remove(envname = env_location,
                             packages = py_pkgs,
                             conda = conda_full_path)

    # Reinstall
    reticulate::conda_install(packages = pkgs_w_versions,
                              envname = env_location,
                              method = 'conda',
                              conda = conda_full_path,
                              python_version = python_version)

    # Reinstall with pip
    reticulate::conda_install(packages = pip_packages,
                              envname = env_location,
                              method = 'conda',
                              conda = conda_full_path,
                              pip = TRUE,
                              python_version = python_version)
  }
  else if(.Platform[['OS.type']] == 'windows'){
    conda_full_path = paste0(partial_conda_path,'/','condabin/conda.bat')

    # Remove all previous installations
    reticulate::conda_remove(envname = env_location,
                             packages = py_pkgs,
                             conda = conda_full_path)

    # Reinstall
    reticulate::conda_install(packages = pkgs_w_versions,
                              envname = env_location,
                              method = 'conda',
                              conda = conda_full_path,
                              python_version = python_version,
                              channel = c('conda-forge', 'vtraag'))

    # Reinstall with pip
    reticulate::conda_install(packages = pip_packages,
                              envname = env_location,
                              method = 'conda',
                              conda = conda_full_path,
                              pip = TRUE,
                              python_version = python_version)

  }
}
```


# Errors on Windows

## Windows 11 OpenSSL Error 

*Please note that this troubleshooting workflow is for a Windows 11 64-bit machine.*

Unfortunately, it is known that installing miniconda on Windows 11 can sometimes result in an [OpenSSL error](https://github.com/drieslab/Giotto/issues/425#issuecomment-1320499840).

It should be noted that Conda developers are aware of this, and that this particular issue does not have to do with the configuration of the Giotto package. One workaround for this error is described below and on the conda repository, [here](https://github.com/conda/conda/issues/8273#issue-409800067).

First, open a terminal and navigate to the folder in which miniconda lives on the machine. To find this, press the Windows Key and search for "Anaconda Prompt". If anaconda3 is installed on the machine, there may be multiple options for this terminal; choose the Anaconda Prompt with "R-MINI~1" in parenthesis.

Output from the following commands will be provided as a comparative guide. Once the terminal is open, identify the Conda environments on the machine by running the following:


```{r, eval=FALSE}
(base) PS > conda info --envs

# conda environments:
#
# base                  *  C:\Users\matto\AppData\Local\R-MINI~1
# giotto_env               C:\Users\matto\AppData\Local\R-MINI~1\envs\giotto_env
#                          C:\Users\matto\AppData\Local\r-miniconda\envs\giotto_env
#                          C:\Users\matto\anaconda3
```


Change your current directory to the R-MINI\~1 subdirectory. Then, navigate to the Library/bin/ subdirectory within. Use the dir command to search bin for two groups of files. These files will be moved to a different directory, and should fix the issue. There may be more than two files per group here, and that is okay. All of the files with these names will be moved, regardless of the extension.


```{r, eval=FALSE}
(base) PS > cd C:\Users\matto\AppData\Local\R-MINI~1
(base) PS > cd .\Library\bin
(base) PS > dir libssl-1_1-x64.*

#    Directory: C:\Users\matto\AppData\Local\r-miniconda\Library\bin


#  Mode                 LastWriteTime         Length Name
#  ----                 -------------         ------ ----
#  -a----         11/4/2022  11:06 AM         686080 libssl-1_1-x64.dll
#  -a----         11/4/2022  11:06 AM        2338816 libssl-1_1-x64.pdbo

(base) PS > dir libcrypto-1_1-x64.*

#    Directory: C:\Users\matto\AppData\Local\r-miniconda\Library\bin


#  Mode                 LastWriteTime         Length Name
#  ----                 -------------         ------ ----
#  -a----         11/4/2022  11:06 AM        3416064 libcrypto-1_1-x64.dll
#  -a----         11/4/2022  11:06 AM       10219520 libcrypto-1_1-x64.pdb
```


Once these files are located, they may be moved to R-MINI\~1/DLLs/, the proper directory for the search path.


```{r, eval=FALSE}
(base) PS > pwd

# Path
# ----
# C:\Users\matth\AppData\Local\r-miniconda\Library\bin

(base) PS > mv libssl-1_1-x64.* ..\..\DLLs\
(base) PS > mv libcrypto-1_1-x64.* ..\..\DLLs\
```


Finally, change to the DLLs directory within R-MINI\~1 and verify that the files now exist there.


```{r, eval=FALSE}
(base) PS > cd C:\Users\matth\AppData\Local\r-miniconda\DLLs\
(base) PS > dir libssl-1_1-x64.*

#     Directory: C:\Users\matth\AppData\Local\r-miniconda\DLLs


#  Mode                 LastWriteTime         Length Name
#  ----                 -------------         ------ ----
#  -a----         11/4/2022  11:07 AM         686080 libssl-1_1-x64.dll
#  -a----         11/4/2022  11:07 AM        2338816 libssl-1_1-x64.pdb

(base) PS > dir libcrypto-1_1-x64.*

#    Directory: C:\Users\matto\AppData\Local\r-miniconda\DLLs


#  Mode                 LastWriteTime         Length Name
#  ----                 -------------         ------ ----
#  -a----         11/4/2022  11:07 AM        3416064 libcrypto-1_1-x64.dll
#  -a----         11/4/2022  11:07 AM       10219520 libcrypto-1_1-x64.pdb
```


Now that these files have moved, this error should disappear. Activate the Giotto environment, and run python within it to test a package import. The OpenSSL error should no longer occur.


```{r, eval=FALSE}
(base) PS > conda info --envs

# conda environments:
#
# base                  *  C:\Users\matto\AppData\Local\R-MINI~1
# giotto_env               C:\Users\matto\AppData\Local\R-MINI~1\envs\giotto_env
#                          C:\Users\matto\AppData\Local\r-miniconda\envs\giotto_env
#                          C:\Users\matto\anaconda3

(base) PS > conda activate giotto_env
(giotto_env) PS > cd C:\Users\matto\AppData\Local\R-MINI~1\envs\giotto_env
(giotto_env) PS > python
Python 3.10.2 | packaged by conda-forge | (main, Mar  8 2022, 15:47:33) [MSC v.1929 64 bit (AMD64)] on win32
Type "help", "copyright", "credits" or "license" for more information.
>>> import pandas, networkx, igraph, leidenalg, community, sklearn
>>>
```


# UnsatisfiableError

This error results from conflicts within the anaconda and miniconda environment. This error presents itself when conflicting versions of conda live on the same machine; conda environments can only be so isolated from each other. To begin the troubleshooting workflow, open a terminal (macOS, Linux) or an Anaconda Prompt (Windows), and identify the environments on the machine. If anaconda3 is installed on the Windows machine, there may be multiple options for this terminal; choose the Anaconda Prompt with "anaconda3" in parenthesis.

**NOTE:** The following commands will be shown as if within an Anaconda Prompt to emphasize the difference for Windows users; for these purposes, the only difference between terminals is the appearance of the message (i.e., (active_env) PS \>) preceding the textual entry. No output will be shown here as differences in OS, environments, and versions will vary.

First, identify the environments on the machine:


```{r, eval=FALSE}
(base) PS > conda info --envs
```


To proceed, any r-miniconda associated environments will be deleted, and the base environment will be updated. If **any** environment is frequently used for other analyses and a python version update is undesirable, it may be preserved by cloning the environment. The original environment, however, will be removed or updated, so ensure that files and workflows associated with this environment are redirected to the new, cloned environment. Ensure the path of the cloned environment is not associated with r-miniconda.

**It is recommended that conda is updated within any cloned environment (see below).**


```{r, eval=FALSE}
(base) PS > conda create --name my_base_clone --clone base
```


Verify that the clone exists to the proper specifications before proceeding by comparing packages and python versions:


```{r, eval=FALSE}
(base) PS > conda info --envs
(base) PS > conda activate my_base_clone
(my_base_clone) PS > conda update conda
(my_base_clone) PS > python -V
(my_base_clone) PS > conda list
(my_base_clone) PS > conda activate base
(base) PS > python -V
(base) PS > conda list 
```


Ensure the base environment is activated. If the r-miniconda environments are still on the machine, remove them. Specify the r-miniconda environments other than giotto_env, as these will be unique to the machine. This may be done at the command line:


```{r, eval=FALSE}
(base) PS > conda env remove --name giotto_env
(base) PS > conda env remove /path/to/r-miniconda/
```


Alternatively, in R, reticulate can uninstall miniconda and remove the associated environments:


```{r, eval=FALSE}
reticulate::miniconda_uninstall()
```


It is advisable to remove any and all environments which are outdated and/or no longer used.

Recall that by default, Giotto installs a miniconda environment with python v3.10.2 for interfacing with R. Older versions of conda in the base environment cannot handle a python version that high in a different environment. Therefore, the recommended troubleshooting method is to update conda and python within the base environment at a minimum. Updating to python v3.8.5 at a minimum is recommended. It is advisable to update conda and python within *each* environment on the machine if feasible.


```{r, eval=FALSE}
(base) PS > conda update conda
(base) PS > conda update python==3.8.5
```


Finally, close the terminal and open the RStudio, VSCode, or an alternative IDE. Running the following should ensure successful installation:


```{r, eval=FALSE}
library(Giotto)
installGiottoEnvironment(force_environment = TRUE, force_miniconda = TRUE)
```


If the issue persists, please post an issue on the [GitHub](https://github.com/drieslab/Giotto/issues).

