---
title: "How to Submit a GitHub Issue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Submit a GitHub Issue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Please read through the information below before submitting your issue to our GitHub page. The link to report an issue can be found at the bottom of the page.

## GitHub Issues Guidelines

### Some Useful Commands

It is helpful if you provide any relevant information about the software and Giotto object you are using. See below for some tips to get this information.

- What version of R are you using?

```{r, eval=FALSE}
R.version.string
```

```{r, eval=FALSE}
R version 4.3.2 (2023-10-31)
```

- What version of Giotto are you using?

```{r, eval=FALSE}
packageVersion("Giotto")
```

```{r, eval=FALSE}
4.0.2
```

- Are you using the Giotto environment or your own python path?

```{r, eval=FALSE}
# Check Giotto environment
Giotto::checkGiottoEnvironment()
```

```{r, eval=FALSE}
giotto environment found at 
/Users/yuanlab/Library/r-miniconda/envs/giotto_env/bin/pythonw
```

- Show sequential processing steps/commands in the Giotto object

```{r, eval=FALSE}
Giotto::showProcessingSteps(my_giotto_object)
```

```{r, eval=FALSE}
Processing steps: 
0_normalize 

 1_subset 

 2_filter 
	 name info:  tag tag 

 3_feat_stats 

 4_cell_stats 

 5_hvf 
	 name info:  hvf HVFplot 

 6_pca 

 7_umap 

 8_tsne 

 9_nn_network 

 10_cluster 
	 name info:  leiden_clus sNN.pca 

 11_delaunay_spatial_network 
	 name info:  Delaunay_network 

 12_spatial_network 
	 name info:  spatial_network 

 13_create_metafeat 
	 name info:  cluster_metagene 

 14_pca 
	 name info:  custom_pca 

 15_umap 
	 name info:  custom_pca custom_umap 

 16_nn_network 
	 name info:  custom_pca custom_NN 

 17_cluster 
	 name info:  custom_leiden custom_NN
```

- Show the structure of the Giotto object

```{r, eval=FALSE}
str(my_giotto_object, max.level = 3)
```

```{r, eval=FALSE}
Formal class 'giotto' [package "Giotto"] with 23 slots
  ..@ expression         :List of 1
  .. ..$ cell:List of 1
  ..@ expression_feat    : chr "rna"
  ..@ spatial_locs       :List of 1
  .. ..$ cell:List of 1
  ..@ spatial_info       : NULL
  ..@ cell_metadata      :List of 1
  .. ..$ cell:List of 1
  ..@ feat_metadata      :List of 1
  .. ..$ cell:List of 1
  ..@ feat_info          : NULL
  ..@ cell_ID            :List of 1
  .. ..$ cell: chr [1:624] "AAAGGGATGTAGCAAG-1" "AAATGGCATGTCTTGT-1" "AAATGGTCAATGTGCC-1" "AAATTAACGGGTAGCT-1" ...
  ..@ feat_ID            :List of 1
  .. ..$ rna: chr [1:634] "Gna12" "Ccnd2" "Btbd17" "Sox9" ...
  ..@ spatial_network    :List of 1
  .. ..$ cell:List of 2
  ..@ spatial_grid       : NULL
  ..@ spatial_enrichment :List of 1
  .. ..$ cell:List of 1
  ..@ dimension_reduction:List of 1
  .. ..$ cells:List of 1
  ..@ nn_network         :List of 1
  .. ..$ cell:List of 1
  ..@ images             : NULL
  ..@ largeImages        :List of 1
  .. ..$ image:Formal class 'giottoLargeImage' [package "Giotto"] with 13 slots
  ..@ parameters         :List of 18
  .. ..$ 0_normalize                : Named chr [1:14] "mini_visium" "raw" "standard" "TRUE" ...
  .. .. ..- attr(*, "names")= chr [1:14] "gobject" "expression_values" "norm_methods" "library_size_norm" ...
  .. ..$ 1_subset                   : Named chr [1:12] "gobject" "spat_unit" "feat_type" "selected_cell_ids" ...
  .. .. ..- attr(*, "names")= chr [1:12] "gobject" "spat_unit" "feat_type" "cell_ids" ...
  .. ..$ 2_filter                   : Named chr [1:13] "mini_visium" "raw" "1" "5" ...
  .. .. ..- attr(*, "names")= chr [1:13] "gobject" "expression_values" "expression_threshold" "feat_det_in_min_cells" ...
  .. ..$ 3_feat_stats               : Named chr [1:4] "mini_visium" "normalized" "0" "TRUE"
  .. .. ..- attr(*, "names")= chr [1:4] "gobject" "expression_values" "detection_threshold" "return_gobject"
  .. ..$ 4_cell_stats               : Named chr [1:4] "mini_visium" "normalized" "0" "TRUE"
  .. .. ..- attr(*, "names")= chr [1:4] "gobject" "expression_values" "detection_threshold" "return_gobject"
  .. ..$ 5_hvf                      : Named chr [1:17] "mini_visium" "normalized" "cov_groups" "FALSE" ...
  .. .. ..- attr(*, "names")= chr [1:17] "gobject" "expression_values" "method" "reverse_log_scale" ...
  .. ..$ 6_pca                      : Named chr [1:15] "mini_visium" "normalized" "cells" "hvf" ...
  .. .. ..- attr(*, "names")= chr [1:15] "gobject" "expression_values" "reduction" "feats_to_use" ...
  .. ..$ 7_umap                     : Named chr [1:17] "mini_visium" "normalized" "cells" "pca" ...
  .. .. ..- attr(*, "names")= chr [1:17] "gobject" "expression_values" "reduction" "dim_reduction_to_use" ...
  .. ..$ 8_tsne                     : Named chr [1:14] "mini_visium" "normalized" "cells" "pca" ...
  .. .. ..- attr(*, "names")= chr [1:14] "gobject" "expression_values" "reduction" "dim_reduction_to_use" ...
  .. ..$ 9_nn_network               : Named chr [1:11] "mini_visium" "sNN" "pca" "1:5" ...
  .. .. ..- attr(*, "names")= chr [1:11] "gobject" "type" "dim_reduction_to_use" "dimensions_to_use" ...
  .. ..$ 10_cluster                 : Named chr [1:11] "mini_visium" "leiden_clus" "sNN" "sNN.pca" ...
  .. .. ..- attr(*, "names")= chr [1:11] "gobject" "name" "nn_network_to_use" "network_name" ...
  .. ..$ 11_delaunay_spatial_network: Named chr [1:4] "dimensions: sdimx and sdimy" "deldir" "auto" "Delaunay_network"
  .. .. ..- attr(*, "names")= chr [1:4] "dimensions used" "method" "maximum distance threshold" "name of spatial network"
  .. ..$ 12_spatial_network         : Named chr [1:4] "10" "all" "400" "spatial_network"
  .. .. ..- attr(*, "names")= chr [1:4] "k neighbours" "dimensions used" "maximum distance threshold" "name of spatial network"
  .. ..$ 13_create_metafeat         : Named chr [1:5] "mini_visium" "normalized" "cluster_genes" "cluster_metagene" ...
  .. .. ..- attr(*, "names")= chr [1:5] "gobject" "expression_values" "feat_clusters" "name" ...
  .. ..$ 14_pca                     : Named chr [1:16] "mini_visium" "normalized" "cells" "custom_pca" ...
  .. .. ..- attr(*, "names")= chr [1:16] "gobject" "expression_values" "reduction" "name" ...
  .. ..$ 15_umap                    : Named chr [1:19] "mini_visium" "normalized" "cells" "pca" ...
  .. .. ..- attr(*, "names")= chr [1:19] "gobject" "expression_values" "reduction" "dim_reduction_to_use" ...
  .. ..$ 16_nn_network              : Named chr [1:13] "mini_visium" "sNN" "pca" "custom_pca" ...
  .. .. ..- attr(*, "names")= chr [1:13] "gobject" "type" "dim_reduction_to_use" "dim_reduction_name" ...
  .. ..$ 17_cluster                 : Named chr [1:11] "mini_visium" "custom_leiden" "sNN" "custom_NN" ...
  .. .. ..- attr(*, "names")= chr [1:11] "gobject" "name" "nn_network_to_use" "network_name" ...
  ..@ instructions       :List of 13
  .. ..$ python_path     : chr "/Users/yuanlab/Library/r-miniconda/envs/giotto_env/bin/pythonw"
  .. ..$ show_plot       : logi TRUE
  .. ..$ return_plot     : logi FALSE
  .. ..$ save_plot       : logi FALSE
  .. ..$ save_dir        : logi NA
  .. ..$ plot_format     : chr "png"
  .. ..$ dpi             : num 300
  .. ..$ units           : chr "in"
  .. ..$ height          : num 9
  .. ..$ width           : num 9
  .. ..$ is_docker       : logi FALSE
  .. ..$ active_spat_unit: chr "cell"
  .. ..$ active_feat_type: chr "rna"
  ..@ offset_file        : NULL
  ..@ OS_platform        : chr "unix"
  ..@ join_info          : NULL
  ..@ multiomics         : NULL
  ..@ h5_file            : NULL
```

### Warning Messages

- Deprecation warning

```{r, eval=FALSE}
my_giotto_object <- createGiottoObject(raw_exprs = "path/to/expression/data",
                                       spatial_locs = "path/to/spatial/data")
```

```{r, eval=FALSE}
Warning message:
In createGiottoObject(raw_exprs = "/Users/username/path/to/data",  :
    raw_exprs argument is deprecated, use expression argument in the future 
```

**Solution** The code has still run and your object has been created, but in the future use the following command:

```{r, eval=FALSE}
my_giotto_object <- createGiottoObject(expression = "path/to/expression/data",
                                       spatial_locs = "path/to/spatial/data")
```

### Errors

- Is this an error from Giotto?

```{r, eval=FALSE}
Error in adjustGiottoMatrix(gobject = my_giotto_object, expression_values = c("normalized"): covariate column name(s) were not found in the cell metadata 
```

**Solution** In this case, you would need to check your cell metadata (`my_giotto_object@cell_metadata`) and make appropriate adjustments.

- Is this an error from an external package?

```{r, eval=FALSE}
Error in methods::as(as.matrix(x[, -1]), "Matrix") : 
no method or default for coercing "matrix" to "Matrix"
```

**Solution** Ensure that the Matrix library is loaded:

```{r, eval=FALSE}
library(Matrix)
```

## Some steps before posting an issue

- Google (or whichever search engine you prefer) the error
- Look up the documentation for the package the error came from

### Required information to provide

- What code/data did you use? (copy-pasted or an example is helpful)
- What did you expect your result to be?
- What was your result?
- If your result was an error message, please see the category on error messages to determine whether the source was Giotto or an external package.

### How to create a reproducible example

For us to be able to help you at the best of our ability, it\'s helpful if you can provide a reproducible example. This just means that anyone who reads your GitHub issue should be able to accurately reproduce your error.

For more information on minimal reproducible examples check out [this link](https://stackoverflow.com/help/minimal-reproducible-example).

And to see what other people recommend, please check out [this forum](https://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example).

### Example of a Good GitHub issue

- Provide a code example
- Please use the GitHub "insert code" function or wrap your code like so:

```{r, eval=FALSE}
`x <- "hello world"

print(x)`
```

*If applicable, provide a screenshot of the issue*

I am having trouble adjusting my Giotto matix. I have inputted the
following code:

```{r, eval=FALSE}
my_giotto_object <- GiottoData::loadGiottoMini("visium")

my_giotto_object <- adjustGiottoMatrix(gobject = my_giotto_object, 
                                       expression_values = c('normalized'),
                                       covariate_columns = c('nr_genes',
                                                             'total_expr'))
```

I expected that my matrix would be appropriately adjusted, but I got the following error:

```{r, eval=FALSE}
Error in adjustGiottoMatrix(gobject = my_giotto_object, expression_values = c("normalized"): covariate column name(s) were not found in the cell metadata 
```

For more context, I am running R version 4.3.2 and Giotto version 4.0.2. I can send you the data I used if necessary.

### Example of a Bad GitHub issue

I am having trouble adjusting my Giotto matrix. My error says that the covariate column name(s) were not found in the cell meta data. What do I do?

Report An Issue <https://github.com/drieslab/Giotto/issues>
