---
title: "How to Submit a GitHub Issue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Submit a GitHub Issue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Please read through the information below before submitting your issue to our GitHub page. The link to report an issue can be found at the bottom of the page.

## GitHub Issues Guidelines

### Some Useful Commands

It is helpful if you provide any relevant information about the software and Giotto object you are using. See below for some tips to get this information.

- What version of R are you using?

```{r, eval=FALSE}
R.version.string
```

```{r, eval=FALSE}
R version 4.0.4 (2021-02-15)
```

- What version of Giotto are you using?

```{r, eval=FALSE}
packageVersion("Giotto")
```

```{r, eval=FALSE}
2.0.0.9044
```

- Are you using the Giotto environment or your own python path?

```{r, eval=FALSE}
# Check Giotto environment
Giotto:::checkGiottoEnvironment()
```

```{r, eval=FALSE}
giotto environment found at 
/Users/natalie_dr/Library/r-miniconda/envs/giotto_env/bin/pythonw 
```

- Show sequential processing steps/commands in the Giotto object

```{r, eval=FALSE}
Giotto:::showProcessingSteps(my_giotto_object)
```

```{r, eval=FALSE}
Processing steps: 
0_normalize 
```

- Show the structure of the Giotto object

```{r, eval=FALSE}
str(my_giotto_object, max.level = 3)
```

```{r, eval=FALSE}
Formal class 'giotto' [package "Giotto"] with 20 slots
..@ expression         :List of 1
.. ..$ rna:List of 3
..@ expression_feat    : chr "rna"
..@ spatial_locs       :List of 1
.. ..$ raw:Classes ‘data.table’ and 'data.frame':   73655 obs. of  4 variables:
.. .. ..- attr(*, ".internal.selfref")=<externalptr> 
..@ spatial_info       : NULL
..@ cell_metadata      :List of 1
.. ..$ rna:Classes ‘data.table’ and 'data.frame':   73655 obs. of  1 variable:
.. .. ..- attr(*, ".internal.selfref")=<externalptr> 
..@ feat_metadata      :List of 1
.. ..$ rna:Classes ‘data.table’ and 'data.frame':   161 obs. of  1 variable:
.. .. ..- attr(*, ".internal.selfref")=<externalptr> 
..@ feat_info          : NULL
..@ cell_ID            : chr [1:73655] "cell_1" "cell_2" "cell_3" "cell_4" ...
..@ feat_ID            :List of 1
.. ..$ rna: chr [1:161] "Ace2" "Adora2a" "Aldh1l1" "Amigo2" ...
..@ spatial_network    : NULL
..@ spatial_grid       : NULL
..@ spatial_enrichment : NULL
..@ dimension_reduction: NULL
..@ nn_network         : NULL
..@ images             : NULL
..@ parameters         :List of 1
.. ..$ 0_normalize: Named chr [1:13] "my_giotto_object" "standard" "TRUE" "6000" ...
.. .. ..- attr(*, "names")= chr [1:13] "gobject" "norm_methods" "library_size_norm" "scalefactor" ...
..@ instructions       :List of 11
.. ..$ python_path: chr "/Users/natalie_dr/Library/r-miniconda/envs/giotto_env/bin/pythonw"
.. ..$ show_plot  : logi TRUE
.. ..$ return_plot: logi TRUE
.. ..$ save_plot  : logi FALSE
.. ..$ save_dir   : chr "/Users/natalie_dr/Desktop/my_giotto/giotto_user_help"
.. ..$ plot_format: chr "png"
.. ..$ dpi        : num 300
.. ..$ units      : chr "in"
.. ..$ height     : num 9
.. ..$ width      : num 9
.. ..$ is_docker  : logi FALSE
..@ offset_file        : NULL
..@ OS_platform        : chr "unix"
..@ join_info          : NULL
```

### Warning Messages

- Deprecation warning

```{r, eval=FALSE}
my_giotto_object <- createGiottoObject(raw_exprs = "path/to/expression/data",
                                   spatial_locs = "path/to/spatial/data")
```

```{r, eval=FALSE}
Warning message:
In createGiottoObject(raw_exprs = "/Users/username/path/to/data",  :
    raw_exprs argument is deprecated, use expression argument in the future 
```

**Solution** The code has still run and your object has been created, but in the future use the following command:

```{r, eval=FALSE}
my_giotto_object <- createGiottoObject(expression = "path/to/expression/data",
                                   spatial_locs = "path/to/spatial/data")
```

### Errors

- Is this an error from Giotto?

```{r, eval=FALSE}
Error in adjustGiottoMatrix(gobject = my_giotto_object, expression_values = c("normalized"): covariate column name(s) were not found in the cell metadata 
```

**Solution** In this case, you would need to check your cell metadata (`my_giotto_object@cell_metadata`) and make appropriate adjustments.

- Is this an error from an external package?

```{r, eval=FALSE}
Error in methods::as(as.matrix(x[, -1]), "Matrix") : 
no method or default for coercing "matrix" to "Matrix"
```

**Solution** Ensure that the Matrix library is loaded:

```{r, eval=FALSE}
library(Matrix)
```

## Some steps before posting an issue

- Google (or whichever search engine you prefer) the error
- Look up the documentation for the package the error came from

### Required information to provide

- What code/data did you use? (copy-pasted or an example is helpful)
- What did you expect your result to be?
- What was your result?
- If your result was an error message, please see the category on error messages to determine whether the source was Giotto or an external package.

### How to create a reproducible example

For us to be able to help you at the best of our ability, it\'s helpful if you can provide a reproducible example. This just means that anyone who reads your GitHub issue should be able to accurately reproduce your error.

For more information on minimal reproducible examples check out [this link](https://stackoverflow.com/help/minimal-reproducible-example).

And to see what other people recommend, please check out [this forum](https://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example).

### Example of a Good GitHub issue

- Provide a code example
- Please use the GitHub "insert code" function or wrap your code like so:

```{r, eval=FALSE}
`x <- "hello world"

print(x)`
```

*If applicable, provide a screenshot of the issue*

I am having trouble adjusting my Giotto matix. I have inputted the
following code:

```{r, eval=FALSE}
my_giotto_object <- adjustGiottoMatrix(gobject = my_giotto_object, 
                                       expression_values = c('normalized'),
                                       covariate_columns = c('nr_genes',
                                                             'total_expr'))
```

I expected that my matrix would be appropriately adjusted, but I got the following error:

```{r, eval=FALSE}
Error in adjustGiottoMatrix(gobject = my_giotto_object, expression_values = c("normalized"): covariate column name(s) were not found in the cell metadata 
```

For more context, I am running R version 4.04 and Giotto version 2.0.0.9044. I can send you the data I used if necessary.

### Example of a Bad GitHub issue

I am having trouble adjusting my Giotto matrix. My error says that the covariate column name(s) were not found in the cell meta data. What do I do?

Report An Issue <https://github.com/drieslab/Giotto/issues>
