---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# 1. Installing Giotto Suite

Giotto Suite is installed via R **but there are required Python modules** which **must** be installed in order for Giotto Suite to function properly. Below are the instructions for both the installation of the Giotto Suite package and required Python environment.

## 1.1 Requirements

- R (\>= 4.0.1)
- Python (\>= 3.6)
- Windows, MacOS or Linux specific installation tools for [Posit](https://support.posit.co/hc/en-us/articles/200486498-Package-Development-Prerequisites) (previously RStudio)

# 2. Installation

## 2.1 Install Giotto Suite


```{r, eval=FALSE}
# Necessary for installation from R
if(!"devtools" %in% installed.packages()) {
  install.packages("devtools")
}

devtools::install_github("drieslab/Giotto@suite_modular")
```


## 2.2 Install the Giotto Python Environment


```{r, eval=FALSE}
library(Giotto)
installGiottoEnvironment()
```


# 3. Try Giotto in the cloud

You can also run analyses in Giotto on Terra.bio. Take a look on the [Terra tutorial]https://drieslab.github.io/Giotto/articles/terra.html).

Encountering errors? Checkout out the [Troubleshooting page](https://drieslab.github.io/Giotto/articles/errorsfaqsandtips.html) for help. Alternatively, post to an issue to the Giotto GitHub page [here](https://github.com/drieslab/Giotto). Please include the version numbers of R, Giotto, and the OS in use at the time of the issue.

# 4. Session Info


```{r, eval=FALSE}
sessionInfo()
```


```{r, eval=FALSE}

    R version 4.2.2 (2022-10-31 ucrt)
    Platform: x86_64-w64-mingw32/x64 (64-bit)
    Running under: Windows 10 x64 (build 22621)

    Matrix products: default

    locale:
    [1] LC_COLLATE=English_United States.utf8 
    [2] LC_CTYPE=English_United States.utf8   
    [3] LC_MONETARY=English_United States.utf8
    [4] LC_NUMERIC=C                          
    [5] LC_TIME=English_United States.utf8    

    attached base packages:
    [1] stats     graphics  grDevices utils     datasets  methods   base     

    other attached packages:
    [1] Giotto_3.2.1

    loaded via a namespace (and not attached):
     [1] Rcpp_1.0.10       pillar_1.9.0      compiler_4.2.2    tools_4.2.2      
     [5] digest_0.6.30     jsonlite_1.8.3    evaluate_0.20     lifecycle_1.0.3  
     [9] tibble_3.2.1      gtable_0.3.3      lattice_0.20-45   pkgconfig_2.0.3  
    [13] rlang_1.1.0       Matrix_1.5-1      cli_3.4.1         rstudioapi_0.14  
    [17] parallel_4.2.2    yaml_2.3.7        xfun_0.38         fastmap_1.1.0    
    [21] terra_1.7-18      dplyr_1.1.1       knitr_1.42        generics_0.1.3   
    [25] vctrs_0.6.1       grid_4.2.2        tidyselect_1.2.0  glue_1.6.2       
    [29] data.table_1.14.6 R6_2.5.1          fansi_1.0.4       rmarkdown_2.21   
    [33] ggplot2_3.4.1     magrittr_2.0.3    scales_1.2.1      codetools_0.2-18 
    [37] htmltools_0.5.4   colorspace_2.1-0  utf8_1.2.3        munsell_0.5.0    

```
