---
title: "intro_to_giotto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro_to_giotto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is an example of a standard visium analysis. For more information and additional examples, visit out Giotto website at http://giottosuite.com

Running time: This vignette takes ~ 25 seconds to run in a local machine with a processor 2.6 GHz 6-Core Intel Core i7, RAM 16 GB, platform x86_64-apple-darwin20, running under macOS Sonoma 14.4.1.

Load package

```{r setup}
library(Giotto)
```

Load a mini visium object

```{r}
g <- GiottoData::loadGiottoMini("visium")
```

Filtering

```{r}
g <- filterGiotto(g)
```

Normalization

```{r}
g <- normalizeGiotto(g)
```

Add statistics

```{r}
g <- addStatistics(g)
```

Find highly variable features

```{r}
g <- calculateHVF(g)
```

Run PCA

```{r}
g <- runPCA(g)
```

Plot PCA

```{r}
plotPCA(g, 
        cell_color = "leiden_clus")
```

Run UMAP

```{r}
g <- runUMAP(g)
```

Plot UMAP

```{r}
plotUMAP(g,
         cell_color = "leiden_clus")
```

Run tSNE

```{r}
g <- runtSNE(g)
```

Plot tSNE

```{r}
plotTSNE(g,
         cell_color = "leiden_clus")
```

Do clustering

```{r}
g <- doLeidenCluster(g)
```

Spatial plot with clusters

```{r}
spatPlot2D(g,
           cell_color = "leiden_clus")
```

Session info

```{r}
sessionInfo()
```

